An efficient composition of bidirectional programs by tupling and lazy updates

### 2020-02-18 paper/2.minbigul.tex
small updates
done: change def of put_m & get_m

 \smallvspace
    $\qtab \getbxinline{bx} : s \to v, \quad \putbxinline{bx} : s * v \to v$
 \smallvspace

@@

@@

\smallvspace
$\tab \putbx{phead \circ phead}{[[1,2,3],[\,],[4,5]]}{100} = [[100,2,3],[\,],[4,5]]$

$\tab \getbx{phead \circ phead}{[[1,2,3],[\,],[4,5]]} = 1$
\smallvspace

>>>>

 \smallvspace
    $\qtab \getbxinline{bx} : s \to v, \quad \putbxinline{bx} : s * v \to s$
 \smallvspace

@@

Also in the semantics of $\putbxinline{Case}$ and $\getbxinline{Case}$, the return statements are used to express clearly the value of functions. Variables $s'$ and $v'$ wrapped in these returns are neccessary for checking the fi conditions.

\subsection{Examples}

@@

$\tab \putbx{phead \circ phead}{[[1,2,3],[\,],[4,5]]}{100} = [[100,2,3],[\,],[4,5]]$

$\tab \getbx{phead \circ phead}{[[1,2,3],[\,],[4,5]]} = 1$

Next let us look at another more complex example using a case combinator:

    $\tab bsnoc = Case \ cond_{sv} \ cond_s \ bx_1 \ bx_2$ where:\\
    $\tabs{3} \ cond_{sv} = \lambda s.\lambda v.(\textnormal{length } v = 1), \ cond_s = \lambda (s_1,s_2).(s_2 = [\ ])$\\
    $\tabs{3} \ bx_1 =  Replace$\\
    $\tabs{3} \ bx_2 = RearrS \ f_1 \ f_2 \ bx_s$ where:\\
        $\tabs{4} f_1 = f_2^{-1} = \lambda (x:y:ys).(y,(x:ys))$, \\
        $\tabs{4} bx_s = RearrV \ g_1 \ g_2 \ bx_v$ where:\\
            $\tabs{5} g_1 = g_2^{-1} = \lambda (v:vs).(v,vs)$\\
            $\tabs{5} bx_v = Replace \times bsnoc$

### 2020-02-18 paper/3.minbigul_m.tex
done: change def of put_m & get_m

The memoizing version, minBiGUL$_m$, needs only two modifications: $get_m$ and $put_m$ (Definitions~\ref{def:putm} and~\ref{def:getm}). The auxiliary function, $get_{mh}$, is used to retrieve the value of an already evaluated $get$ from the table or to compute an unevaluated $get$ by $get_m$ and store the key-value tuple in the table (Definition~\ref{def:getmh}).

@@

    \noindent $\putm{bx}{s}{v} = \match{bx}\\
        \tab \vert \ bx_1 \circ bx_2 \to \putm{bx_1}{s}{(\putm{bx_2}{(\getmh{bx_1}{s})}{v})} \\
        \tab \vert \ \_ \to \text{similar to } put$
\end{definition}

\begin{multicols}{2}
    \begin{definition} \label{def:getm} $\text{Memo version of } get$

        \noindent $\getm{bx}{s} = \match{bx}\\
            \tab \ \vert \ bx_1 \circ bx_2 \to\\
            \tabs{2} \textnormal{try } (\text{Hashtbl.find} \ table_g \ (bx, s))\\
            \tabs{2} \textnormal{with } \text{Not\_found } \to\\
                \tabs{3} v \Leftarrow \getmh{bx_2}{(\getmh{bx_1}{s})}\\
                \tabs{3} \text{Hashtbl.add } \ table_g \ (bx, s) \ v\\
                \tabs{3} v \\
            \tab \ \vert \ \_ \to \text{similar to } get$
    \end{definition}
\columnbreak
    \begin{definition} \label{def:getmh} $\text{Interface for } get_m$

        \noindent $\getmh{bx}{s} =\\
            \tab \textnormal{try } (\text{Hashtbl.find} \ table_g \ (bx, s))\\
            \tab \textnormal{with } \text{Not\_found } \to\\
                \tabs{2} v \Leftarrow \getm{bx}{s}\\
                \tabs{2} \match{bx} \\
                \tabs{2} \ \vert \ bx_1 \circ bx_2 \to v \\
                \tabs{2} \ \vert \ \_ \to \text{Hashtbl.add } \ table_g \ (bx, s) \ v\\
                \qtabs{2} \ v$
    \end{definition}
\end{multicols}

% Note as a special point 
% It should be noted 
Note in particular 
that the interpreter does not save all states when evaluating a program, only the intermediate states of a composition. When evaluating $\putm{bx_1 \circ bx_2}{s}{v}$, the function $get_{mh}$ is used, not $get$ or $get_m$.
% appears to indicate the saving. 
When evaluating $\getm{bx_1 \circ bx_2}{s}$, the key ($bx,s$), where $bx \equiv bx_1 \circ bx_2$, is looked up in the table and the corresponding value is used for the next steps in the evaluation. If there is no such key, the value $v$ will be calculated by using $get_{mh}$. This key-value insertion is necessary because the interpreter may later need to reevaluate $\getm{bx_1 \circ bx_2}{s}$.
% reevaluate <--> evaluate one more time.

>>>>

The memoizing version, minBiGUL$_m$, needs only two modifications: $get_m$ and $put_m$ (Definitions~\ref{def:putm} and~\ref{def:getm}).

@@

    \noindent $\putm{bx}{s}{v} = \match{bx}\\
        \tab \vert \ bx_1 \circ bx_2 \to \\
            \tabs{2} \putm{bx_1}{s}{(\putm{bx_2}{(\getm{bx_1}{s})}{v})} \\
        \tab \vert \ \_ \to \text{similar to } put$
\end{definition}

\begin{definition} \label{def:getm} $\text{Memo version of } get$

    \noindent $\getm{bx}{s} = \match{bx}\\
    \tab \ \vert \ bx_1 \circ bx_2 \to\\
    \tabs{2} \textnormal{try } (\text{Hashtbl.find} \ table_g \ (bx, s))\\
    \tabs{2} \textnormal{with } \text{Not\_found } \to\\
        \tabs{3} i \Leftarrow \getm{bx_1}{s}; \tab \text{Hashtbl.add } \ table_g \ (bx_1, s) \ i;\\
        \tabs{3} v \Leftarrow \getm{bx_2}{i}; \tab \text{Hashtbl.add } \ table_g \ (bx_2, i) \ v;\\
        \tabs{3} \text{Hashtbl.add } \ table_g \ (bx, s) \ v\\
        \tabs{3} v \\
    \tab \ \vert \ \_ \to \text{similar to } get$
\end{definition}

The evaluation of $\putm{bx_1 \circ bx_2}{s}{v}$ includes two recursive calls of $put_m$ and an external call of $get_m$, which is relatively similar to the evaluation of $\putbx{bx_1 \circ bx_2}{s}{v}$. 
Meanwhile, the evaluation of $\getm{bx_1 \circ bx_2}{s}$ does not merely invoke $get_m$ recursively twice. In case that $bx$ is a composition, the key ($bx,s$) needs to be looked up in the table and the corresponding value would be used for the next steps in the evaluation. If there is no such key, the value of the intermediate state $i$ and the value of $\getbx{bx}{s}$ in $v$ will be calculated. These values along with the corresponding keys will also be stored in the table where the interpreter may later 
leverage instead of reevaluating some states.
% Note as a special point 
% It should be noted 
Note in particular 
that the interpreter does not save all states when evaluating a program, only the intermediate states of a composition.

### 2020-02-18 paper/4.xpg.tex
done: change def of put_m & get_m

Constructions of $pg$ for the replacement, the product and the source/view rearrangements are very clear when just pairing $put$ and $get$ respectively, then doing basic changes.\\

    \noindent $\tabs{2} \pg{\casebx{cond_{sv}}{cond_{s}}{bx_1}{bx_2}}{s}{v} \\
    \tabs{3} \overset{1}{=} (\text{if} \ {cond_{sv} \ s \ v} \qtabs{3} \tab \text{if} \ {cond_{s} \ s} \\
    \tabs{2} \qtab \text{then} \ s' \Leftarrow \putbx{bx_1}{s}{v} \qtab \tab \, \text{then} \ v' \Leftarrow \getbx{bx_1}{s} \\
    \tabs{2} \qtab \text{else} \ s' \Leftarrow \putbx{bx_2}{s}{v} \qtab \tab \ \, \text{else} \ v' \Leftarrow \getbx{bx_2}{s}\\
    \tabs{2} \qtab \text{fi} \ cond_{s} \ s'; \text{ return } s' \qtab , \tab \ \ \text{fi} \ {cond_{sv} \ s \ v'}; \text{ return } v')\\
    \tabs{3} \overset{2}{=} \text{if} \ {cond_{sv} \ s \ v} \ \&\& \ {cond_{s} \ s}\\
    \tabs{2} \qtab \text{then} \ (s', v') \Leftarrow \pg{bx_1}{s}{v}\\
    \tabs{2} \qtab \text{else} \ (s', v') \Leftarrow \pg{bx_2}{s}{v}\\
    \tabs{2} \qtab \text{fi} \ cond_{s} \ s' \ \&\& \ cond_{sv} \ s \ v'; \text{ return } (s',v')$

@@

    \noindent $\tabs{2} \pg{bx_1 \circ bx_2}{s}{v} \\
    \tabs{3} \overset{1}{=} (\putbx{bx_1}{s}{(\putbx{bx_2}{(\getbx{bx_1}{s})}{v})}, \getbx{bx_2}{(\getbx{bx_1}{s})}) \\
    \tabs{3} \overset{2}{=} v_1 \Leftarrow \getbx{bx_1}{s}; \qtabs{3} \ \, \overset{3}{=} (s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy};\\
        \tabs{2} \qtab (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v}; \qtabs{2} (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v};\\
        \tabs{2} \qtab (s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}; \qtabs{2} (s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2};\\
            \qtab \tabs{3} (s_3, v_2) \qtabs{6} (s_3, v_2)\\
    \tabs{3} \overset{4}{=} (s_1, v_1) \Leftarrow \pg{bx_1}{s}{construct\_dummy \ s}; \\
        \tabs{2} \qtab (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v}; \\
        \tabs{2} \qtab (s_3, v_3) \Leftarrow \pg{bx_1}{s_1}{s_2}; \\
            \qtab \tabs{3} (s_3, v_2)$

The construction of $\pginline{bx_1 \circ bx_2}$ can be considered as the soul of the $pg$ function. The first two equalities comes from mentioned definitions and some basic transformations. The third one rewrites $v_1 \Leftarrow \getbx{bx_1}{s}$ into $(s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy}$. This is possible when we consider $\getbx{bx_1}{s}$ as the second element of $\pg{bx_1}{s}{dummy}$ where $dummy$ is a special value that makes the $put \, [\![bx_1]\!]$ valid.
%We can construct a $dummy$ from the source by \textsc{GetPut} law in some cases or require programmers to give a valid view.
Since there is no real view, this $dummy$ is necessary. The last equality changes two. First is a change from $dummy$ to an application $construct\_dummy \ s$. This means we need a way to construct $dummy$ from the current source, $s$. We can require programmers to give this $construct\_dummy$ function. Second is a substitution under the \textsc{PutPut} law on $s$ of $(s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}$ by $s_1$ which equals to $\putbx{bx_1}{s}{(construct\_dummy \ s)}$. $v_3$ in the result pair $(s_3,v_3)$ equals $dummy$. So we realize that there is no loss information when computing a composition.

>>>>

Constructions of $pg$ for the replacement, the product and the source/view rearrangements are very clear when just pairing $put$ and $get$ respectively, then doing basic changes.
%
The values of these $pg$ functions are the values of final expressions respectively. We only use the return keyword to express explicitly the evaluated value of a function in the situation of $Case$.

    \noindent $\tabs{2} \pg{\casebx{cond_{sv}}{cond_{s}}{bx_1}{bx_2}}{s}{v} \\
    \tabs{3} \overset{1}{=} (\text{if} \ {cond_{sv} \ s \ v} \qtabs{3} \tab \text{if} \ {cond_{s} \ s} \\
    \tabs{2} \qtab \text{then} \ s' \Leftarrow \putbx{bx_1}{s}{v} \qtab \tab \, \text{then} \ v' \Leftarrow \getbx{bx_1}{s} \\
    \tabs{2} \qtab \text{else} \ s' \Leftarrow \putbx{bx_2}{s}{v} \qtab \tab \ \, \text{else} \ v' \Leftarrow \getbx{bx_2}{s}\\
    \tabs{2} \qtab \text{fi} \ cond_{s} \ s'; \text{ return } s' \qtab , \tab \ \ \text{fi} \ {cond_{sv} \ s \ v'}; \text{ return } v')\\
    %
    \tabs{3} \overset{2}{=} \text{if} \ {cond_{sv} \ s \ v} \ \&\& \ {cond_{s} \ s} \ \text{then}\\
        \qtabs{2} \tab (s', v') \Leftarrow (\putbx{bx_1}{s}{v}, \getbx{bx_1}{s});\\
        \qtabs{2} \tab \text{if} \ cond_{s} \ s' \ \&\& \ cond_{sv} \ s \ v' \ \text{then return } (s',v') \ \text{else } \textit{undefined}\\
        \tabs{2} \qtab \text{else if} \ {cond_{sv} \ s \ v} \ \&\& \ not \ ({cond_{s} \ s}) \ \text{then}\\
        \qtabs{2} \tab (s', v') \Leftarrow (\putbx{bx_1}{s}{v}, \getbx{bx_2}{s});\\
        \qtabs{2} \tab \text{if} \ cond_{s} \ s' \ \&\& \ not \ (cond_{sv} \ s \ v') \ \text{then return } (s',v') \ \text{else } \textit{undefined}\\
        \tabs{2} \qtab \text{else if} \ not \ ({cond_{sv} \ s \ v}) \ \&\& \ {cond_{s} \ s} \ \text{then}\\
        \qtabs{2} \tab (s', v') \Leftarrow (\putbx{bx_2}{s}{v}, \getbx{bx_1}{s});\\
        \qtabs{2} \tab \text{if} \ not \ (cond_{s} \ s') \ \&\& \ cond_{sv} \ s \ v' \ \text{then return } (s',v') \ \text{else } \textit{undefined}\\
        \tabs{2} \qtab \text{else if} \ not \ ({cond_{sv} \ s \ v}) \ \&\& \ not \ ({cond_{s} \ s}) \ \text{then}\\
        \qtabs{2} \tab (s', v') \Leftarrow (\putbx{bx_2}{s}{v}, \getbx{bx_2}{s});\\
        \qtabs{2} \tab \text{if} \ not \ (cond_{s} \ s') \ \&\& \ not \ (cond_{sv} \ s \ v') \ \text{then return } (s',v') \ \text{else } \textit{undefined}\\
    %
    \tabs{3} \overset{3}{=} (* \text{ with restriction } *)\\
    \tabs{2} \qtab \text{if} \ {cond_{sv} \ s \ v} \ \&\& \ {cond_{s} \ s}\\
    \tabs{2} \qtab \text{then} \ (s', v') \Leftarrow \pg{bx_1}{s}{v}\\
    \tabs{2} \qtab \text{else} \ (s', v') \Leftarrow \pg{bx_2}{s}{v}\\
    \tabs{2} \qtab \text{fi} \ cond_{s} \ s' \ \&\& \ cond_{sv} \ s \ v'; \text{ return } (s',v')$

@@

    \noindent $\tabs{2} \pg{bx_1 \circ bx_2}{s}{v} \\
    \tabs{3} \overset{1}{=} (\putbx{bx_1}{s}{(\putbx{bx_2}{(\getbx{bx_1}{s})}{v})}, \getbx{bx_2}{(\getbx{bx_1}{s})}) \\
    \tabs{3} \overset{2}{=} v_1 \Leftarrow \getbx{bx_1}{s}; \qtabs{3} \ \, \overset{3}{=} (s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy};\\
        \tabs{2} \qtab (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v}; \qtabs{2} (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v};\\
        \tabs{2} \qtab (s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}; \qtabs{2} (s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2};\\
            \qtab \tabs{3} (s_3, v_2) \qtabs{6} (s_3, v_2)\\
    \tabs{3} \overset{4}{=} (s_1, v_1) \Leftarrow \pg{bx_1}{s}{construct\_dummy \ s}; \\
        \tabs{2} \qtab (s_2, v_2) \Leftarrow \pg{bx_2}{v_1}{v}; \\
        \tabs{2} \qtab (s_3, v_3') \Leftarrow \pg{bx_1}{s_1}{s_2}; \\
            \qtab \tabs{3} (s_3, v_2)$

The construction of $\pginline{bx_1 \circ bx_2}$ can be considered as the soul of the $pg$ function. The first two equalities comes from mentioned definitions and some basic transformations. The third one rewrites $v_1 \Leftarrow \getbx{bx_1}{s}$ into $(s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy}$. This is possible when we consider $\getbx{bx_1}{s}$ as the second element of $\pg{bx_1}{s}{dummy}$ where $dummy$ is a special value that makes the $put \, [\![bx_1]\!]$ valid.
%We can construct a $dummy$ from the source by \textsc{GetPut} law in some cases or require programmers to give a valid view.
Since there is no real view, this $dummy$ is necessary to pair with original source $s$ to form the input of $put \, [\![bx_1]\!]$. The last equality changes two. The first is a change from $dummy$ to an application $construct\_dummy \ s$. This means we need a way to construct $dummy$ from the current source, $s$. Programmers can be required to give this $construct\_dummy$ function, but it may be inessential for ill-typed systems where choosing $dummy$ as $v$ is the easiest way to meet our expectation. We set up that choice in our experiments but still use the $construct\_dummy$ function in the later definitions to provide a more general meanning.
%
The second is the change from $(s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}$ to $(s_3, v_3') \Leftarrow \pg{bx_1}{s_1}{s_2}$, where $s$ and $v_3$ are replaced with $s_1$ and $v_3'$ respectively. Because $s_1$ is a source update of $\putbx{bx_1}{s}{dummy}$, so under the \textsc{PutPut} law, it is possible to substitute $s$ by $s_1$ is possible. The substitution of $v_3$ by $v_3'$ is simply replacing the variable name since $v_3$ and $v_3'$ hold different results of $\getbx{bx_1}{s}$ and $\getbx{bx_1}{s_1}$ respectively. Because both variables are no longer used later, this substitution does not affect the outcome of the function.
